A Bayesian approach to uncertainty in word embedding bias estimation

### 2023-06-22 paper/CLsubmission/bayesBiasCL.tex
CLsubmission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  dvipsnames,enabledeprecatedfontcommands]{scrartcl}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
%\documentclass{article}

% %packages
\usepackage{booktabs}
%\usepackage[left]{showlabels}
\usepackage{multirow}
\usepackage{subcaption}

\usepackage{graphicx}
%\usepackage{longtable}
\usepackage{supertabular, booktabs}
\usepackage{ragged2e}
\usepackage{etex}
%\usepackage{yfonts}
\usepackage{marvosym}
\usepackage[notextcomp]{kpfonts}
\usepackage{nicefrac}
\newcommand*{\QED}{\hfill \footnotesize {\sc Q.e.d.}}
\usepackage{floatrow}

\usepackage[textsize=footnotesize]{todonotes}

%\linespread{1.5}
\newcommand{\indep}{\!\perp \!\!\! \perp\!}

\setlength{\parindent}{5mm}
\setlength{\parskip}{1pt}

%language
%\usepackage{times}
\usepackage{mathptmx}
\usepackage[scaled=0.86]{helvet}
\usepackage{t1enc}
%\usepackage[utf8x]{inputenc}
%\usepackage[polish]{babel}
%\usepackage{polski}

%AMS
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{mathtools}

\usepackage{geometry}
 \geometry{a4paper,left=35mm,top=20mm,}

%environments
\newtheorem{fact}{Fact}

%abbreviations
\newcommand{\ra}{\rangle}
\newcommand{\la}{\langle}
\newcommand{\n}{\neg}
\newcommand{\et}{\wedge}
\newcommand{\jt}{\rightarrow}
\newcommand{\ko}[1]{\forall  #1\,}
\newcommand{\ro}{\leftrightarrow}
\newcommand{\exi}[1]{\exists\, {_{#1}}}
\newcommand{\pr}[1]{\ensuremath{\mathsf{P}(#1)}}
\newcommand{\cost}{\mathsf{cost}}
\newcommand{\benefit}{\mathsf{benefit}}
\newcommand{\ut}{\mathsf{ut}}

\newcommand{\odds}{\mathsf{Odds}}
\newcommand{\ind}{\mathsf{Ind}}
\newcommand{\nf}[2]{\nicefrac{#1\,}{#2}}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\prr}[1]{\mbox{$\mathtt{P}_{prior}(#1)$}}
\newcommand{\prp}[1]{\mbox{$\mathtt{P}_{posterior}(#1)$}}

\newtheorem{q}{\color{blue}Question}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}[fact]

%technical intermezzo
%---------------------

\newcommand{\intermezzoa}{
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Starts}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}\nopagebreak 
	}

\newcommand{\intermezzob}{\nopagebreak 
	\begin{minipage}[c]{13cm}
	\begin{center}\rule{10cm}{0.4pt}

	\tiny{\sc Optional Content Ends}

	\vspace{-1mm}

	\rule{10cm}{0.4pt}\end{center}
	\end{minipage}
	}

%--------------------

\newtheorem*{reply*}{Reply}
\usepackage{enumitem}
\newcommand{\question}[1]{\begin{enumerate}[resume,leftmargin=0cm,labelsep=0cm,align=left]
\item #1
\end{enumerate}}

\usepackage{float}

% \usepackage{sectsty}
% \sectionfont{\fontsize{12}{11}\selectfont}
% \subsectionfont{\fontsize{11}{11}\selectfont}

% \setbeamertemplate{blocks}[rounded][shadow=true]
% \setbeamertemplate{itemize items}[ball]
% \AtBeginPart{}
% \AtBeginSection{}
% \AtBeginSubsection{}
% \AtBeginSubsubsection{}
% \setlength{\emergencystretch}{0em}
% \setlength{\parskip}{0pt}

\usepackage[authoryear]{natbib}

%\bibliographystyle{apalike}

\usepackage{tikz}
\usetikzlibrary{positioning,shapes,arrows}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Bayesian approach to uncertainty in word embedding bias estimation},
  pdfauthor={Alicja Dobrzeniecka and Rafal Urbaniak},
  colorlinks=true,
  linkcolor={Maroon},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Bayesian approach to uncertainty in word embedding bias
estimation}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

\begin{document}
\maketitle

\textbf{Abstract.} Multiple measures, such as WEAT or MAC, attempt to
quantify the magnitude of bias present in word embeddings in terms of a
single-number metric. However, such metrics and the related statistical
significance calculations rely on treating pre-averaged data as
individual data points and employing bootstrapping techniques with low
sample sizes. We show that similar results can be easily obtained using
such methods even if the data are generated by a null model lacking the
intended bias. Consequently, we argue that this approach generates false
confidence. To address this issue, we propose a Bayesian alternative:
hierarchical Bayesian modeling, which enables a more
uncertainty-sensitive inspection of bias in word embeddings at different
levels of granularity. To showcase our method, we apply it to
\emph{Religion, Gender}, and \emph{Race} word lists from the original
research, together with our control neutral word lists. We deploy the
method using \emph{Google, Glove}, and \emph{Reddit} embeddings.
Further, we utilize our approach to evaluate a debiasing technique
applied to the \emph{Reddit} word embedding. Our findings reveal a more
complex landscape than suggested by the proponents of single-number
metrics. The datasets and source code for the paper are publicly
available.\footnote{\url{https://github.com/efemeryds/Bayesian-analysis-for-NLP-bias}}

@@

It has been suggested\footnote{See for instance {[}1,4,5,9,12,13{]}.}
that language models can learn implicit biases that reflect harmful
stereotypical thinking---for example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity {[}16{]}.

One response to the raising of the issue of bias in natural language
models might be to say that there is not much point to reflecting on
such biases, as they are unavoidable. This unavoidability might seem in
line with the arguments to the effect that learning algorithms are
always value-laden {[}10{]}: they employ inductive methods that require
design-, data-, or risk-related decisions that have to be guided by
extra-algorithmic considerations. Such choices necessarily involve value
judgments and have to do, for instance, with what simplifications or
risks one finds acceptable. Admittedly, algorithmic decision making
cannot fulfill the value-free ideal, but this only means that even more
attention needs to be paid to the values underlying different techniques
and decisions, and to the values being pursued in a particular use of an
algorithm.

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here, since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, which is shaped in
part by the natural language processing tools we use on a massive scale.
Sure, if there is unfairness and our goal is to diagnose it, we should
do complete justice to learning it in the model used to study it. One
example of this approach is {[}3{]}, where the authors use language
models to study the shape of certain biases across a century.

However, if our goal is to develop downstream tools that perform tasks
that we care about without further perpetuating or exacerbating harmful
stereotypes, we still have good reasons to try to minimize the negative
impact. Moreover, it is often not the case that the corpora mirror
reality---to give a trivial example, heads are spoken of more often than
kidneys, but this does not mean that kidneys occur much less often in
reality than heads. To give a more relevant example, the
disproportionate association of female words with female occupations in
a corpus actually greatly exaggerates the actual lower disproportion in
the real distribution of occupations {[}6{]}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) {[}9{]}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) {[}13{]}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for {[}9{]} the range for protected words is between 13 and
  100, and for attributes between 16 and 25; for {[}13{]} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

@@

The underlying intuition is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history {[}3{]},
  embeddings which do not protect certain classes may also be useful.}

@@

One of the first measures of bias has been developed in {[}1{]}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in {[}1{]}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(w\) is then understood as \(w\)'s projection
on the gender direction: \(\vec{w} \cdot gd\) (which, after normalizing
by dividing by \(\lVert w \rVert \,\lVert \mathsf{gd} \rVert\), is the
same as cosine similarity). Given a list \(N\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. See our list in Appendix \ref{app:custom} and a brief
  discussion in Subsection \ref{subsec:interpretability}.} and the
gender direction \(\mathsf{gd}\), the direct gender bias is defined as
the average cosine similarity of the words in \(N\) from \(\mathsf{gd}\)
(\(c\) is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in {[}5{]}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in {[}5{]} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In {[}1{]} another method which involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  {[}18{]}.}

A measure of bias in word embeddings which does not proceed by
identifying bias directions (such as a gender vector), the Word
Embedding Association Test (\textsf{WEAT}), has been proposed in
{[}9{]}. The idea here is that the bias between two sets of target
words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we will call them attributes). For instance, \(X\)
might be a set of male names, \(Y\) a set of female names, \(A\) might
contain stereotypically male-related, and \(B\) stereotypically
female-related career words. The association difference for a particular
word \(t\) (belonging to either \(X\) or \(Y\)) is:

@@

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
{[}19{]} used in psychology, and in some applications it uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. In {[}9{]} the authors argue that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. In {[}12{]} the methodology is extended
to a multilingual and cross-lingual setting, arguing that using
Euclidean distance instead of cosine similarity does not make much
difference, while the bias effects vary greatly across embedding
models.\footnote{Interestingly, with social media-text trained
  embeddings being less biased than those based on Wikipedia.} A similar
methodology is employed in {[}4{]}. The authors employ word embeddings
trained on corpora from different decades to study the shifts in various
biases through the century.\footnote{Strictly speaking, these authors
  use Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

\begin{figure}[H]

\begin{center}\includegraphics[width=0.55\linewidth]{paperDraft_files/figure-latex/WEATexample5-1} \end{center}
\caption{A simple example of a two-class set-up. Two groups of protected words, $X$ and $Y$ with two stereotypical attribute sets. An example of \textsf{WEAT} calculations follows.}
\label{eq:WEATexample}
\end{figure}

@@

\textsf{WEAT} has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, so the question arises as to
how to generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in {[}13{]}. The authors introduce Mean
Average Cosine distance (\textsf{MAC}) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\begin{figure}[H]

\begin{center}\includegraphics[width=0.5\linewidth]{paperDraft_files/figure-latex/MACexample-1} \end{center}
\caption{A small subset of the religion dataset.  To each protected word in $T$  there corresponds one class of stereotypical attributes typically associated with it (and other classes of stereotypical attributes associated with different protected words).}
\label{fig:MACexample}
\end{figure}

@@

Notably, the inutitive distinction between different attribute sets
plays no real role in the \textsf{MAC} calculations. Equally well one
could calculate the mean distance of \emph{muslim} to all the
predicates, mean distance of \emph{christian} to all the predicates,
mean distance of \emph{jew} to all the predicates, and then to take the
mean of these three means.

@@

\begin{figure}[H]

\begin{center}\includegraphics[width=0.7\linewidth]{paperDraft_files/figure-latex/fig:cosineDistributions8d-1} \end{center}

@@

\begin{figure}[H]

\begin{center}\includegraphics[width=0.7\linewidth]{paperDraft_files/figure-latex/fig:cosineDistributions8df-1} \end{center}

@@

\begin{figure}[H]

>>>>

\documentclass{clv3}

\usepackage{amsmath,amssymb}
\usepackage{float}
\usepackage{nicefrac}

\usepackage{graphicx}

\usepackage{booktabs}
%\usepackage[left]{showlabels}
\usepackage{multirow}
\usepackage{subcaption}

\usepackage{url}
\usepackage{hyperref}
\usepackage[table]{xcolor}
\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true,citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}

\bibliographystyle{compling}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\begin{document}

%\issue{1}{1}{2016}
%Document Head
%\dochead{CLV3 Class File Manual}

\runningtitle{Bayesian word embedding bias estimation}

\runningauthor{Dobrzeniecka A. and Urbaniak, R.}

%\pageonefooter{Action editor: \{action editor name\}. Submission received: DD Month YYYY; revised version received: DD Month YYYY; accepted for publication: DD Month YYYY.}

\title{A Bayesian approach to uncertainty in word embedding bias
estimation}

\author{Alicja Dobrzeniecka}

\affil{Vrije Universiteit Amsterdam \\ The Netherlands}

\author{Rafal Urbaniak%\thanks{PITC Building}
}
\affil{Basis Research Institute, NYC, USA  \\  The University of Gdansk,  Poland}

%\author{And Another Author}
%\affil{Publishing / SPi}

%author{And Yet Another}
%\affil{Publishing / SPi}

\maketitle

\begin{abstract}
Multiple measures, such as WEAT or MAC, attempt to
quantify the magnitude of bias present in word embeddings in terms of a
single-number metric. However, such metrics and the related statistical
significance calculations rely on treating pre-averaged data as
individual data points and employing bootstrapping techniques with low
sample sizes. We show that similar results can be easily obtained using
such methods even if the data are generated by a null model lacking the
intended bias. Consequently, we argue that this approach generates false
confidence. To address this issue, we propose a Bayesian alternative:
hierarchical Bayesian modeling, which enables a more
uncertainty-sensitive inspection of bias in word embeddings at different
levels of granularity. To showcase our method, we apply it to
Religion, Gender, and Race word lists from the original
research, together with our control neutral word lists. We deploy the
method using Google, Glove, and Reddit embeddings.
Further, we utilize our approach to evaluate a debiasing technique
applied to the Reddit word embedding. Our findings reveal a more
complex landscape than suggested by the proponents of single-number
metrics. The datasets and source code for the paper are publicly
available.\footnote{The datasets and source code for the paper are publicly
available at \url{https://github.com/efemeryds/Bayesian-analysis-for-NLP-bias}.}
\end{abstract}

@@

It has been suggested\footnote{See for instance \citep{Bolukbasi2016man, Caliskan2017semanticsBiases, Gonen2019lipstick, Lauscher2019multidimensional, Garg2018years, Manzini2019blackToCriminal}.}
that language models can learn implicit biases that reflect harmful
stereotypical thinking---for example, the (vector corresponding to the)
word \textit{she} might be much closer in the vector space to the word
\textit{cooking} than the word \textit{he}. Such phenomena are
undesirable at least in some downstream tasks, such as web search,
recommendations, and so on. To investigate such issues, several measures
of bias in word embeddings have been formulated and applied. Our goal is
to use two prominent examples of such measures to argue that this
approach oversimplifies the situation and to develop a Bayesian
alternative.

A common approach in natural language processing is to represent words
by vectors of real numbers---such representations are called
\emph{embeddings}. One way to construct an embedding---we will focus our
attention on non-contextual language models\footnote{One example of a
  contextualized representation is BERT. Another is GPT.}---is to use a
large corpus to train a neural network to assign vectors to words in a
way that optimizes for co-occurrence prediction accuracy. Such vectors
can then be compared in terms of their similarity---the usual measure is
cosine similarity---and the results of such comparisons can be used in
downstream tasks. Roughly speaking, cosine similarity is an imperfect
mathematical proxy for semantic similarity \citep{Mikolov2013efficient}.

One response to the raising of the issue of bias in natural language
models might be to say that there is not much point in reflecting on
such biases, as they are unavoidable. This unavoidability might seem in
line with the arguments to the effect that learning algorithms are
always value-laden \citep{JohnsonValueFree}: they employ inductive methods that require
design-, data-, or risk-related decisions that have to be guided by
extra-algorithmic considerations. Such choices necessarily involve value
judgments and have to do, for instance, with what simplifications or
risks one finds acceptable. Admittedly, algorithmic decision making
cannot fulfill the value-free ideal, but this only means that even more
attention needs to be paid to the values underlying different techniques
and decisions, and to the values being pursued in a particular use of an
algorithm.

Another response might be to insist that there is no bias introduced by
the use of machine learning methods here since the algorithm is simply
learning to correctly predict co-occurrences based on what ``reality''
looks like. However, this objection overlooks the fact that we, humans,
are the ones who construct this linguistic reality, which is shaped in
part by the natural language processing tools we use on a massive scale.
Sure, if there is unfairness and our goal is to diagnose it, we should
do complete justice to learning it in the model used to study it. One
example of this approach is \citep{Garg2017hundredYears}, where the authors use language
models to study the shape of certain biases across a century.

However, if our goal is to develop downstream tools that perform tasks
that we care about without further perpetuating or exacerbating harmful
stereotypes, we still have good reasons to try to minimize the negative
impact. Moreover, it is often not the case that the corpora mirror
reality---to give a trivial example, heads are spoken of more often than
kidneys, but this does not mean that kidneys occur much less often in
reality than heads. To give a more relevant example, the
disproportionate association of female words with female occupations in
a corpus actually greatly exaggerates the actual lower disproportion in
the real distribution of occupations \citep{gordon2012reporting}.

In what follows, we focus on two popular measures of bias applicable to
many existing word embeddings, such as
\emph{GoogleNews},\footnote{GoogleNews-vectors-negative300, available at  \url{https://github.com/mmihaltz/word2vec-GoogleNews-vectors}.}
\emph{Glove}\footnote{Available at \url{https://nlp.stanford.edu/projects/glove/}.}
and
\emph{Reddit Corpus}\footnote{Reddit-L2 corpus, available at  \url{http://cl.haifa.ac.il/projects/L2/}.}:
\emph{Word Embedding Association Test} (\textsf{WEAT}) \citep{Caliskan2017semanticsBiases}, and
\emph{Mean Average Cosine Distance} (\textsf{MAC}) \citep{Manzini2019blackToCriminal}. We first
explain how these measures are supposed to work. Then we argue that they
are problematic for various reasons---the key one being that by
pre-averaging data they manufacture false confidence, which we
illustrate in terms of simulations showing that the measures often
suggest the existence of bias even if by design it is non-existent in a
simulated dataset.

@@

Most of the problems that we point out generalize to any existing
approach that focuses on chasing a single numeric metric of bias. (1)
They treat the results of pre-averaging as raw data in statistical
significance tests, which in this context is bound to overestimate
significance. We show similar results can easily be obtained when
sampling from null models with no bias. (2) The word list sizes and
sample sizes used in the studies are usually small,\footnote{Depending
  on a list for \citep{Caliskan2017semanticsBiases} the range for protected words is between 13 and
  100, and for attributes between 16 and 25; for \citep{Manzini2019blackToCriminal} the range for
  protected words is between 14 and 18, and for attributes between 11
  and 25.} (3) Many studies do not use any control predicates, such as
random neutral words or neutral human predicates for comparison.

@@

The underlying intuition is that if a particular harmful stereotype is
learned in a particular embedding, then certain groups of words will be
systematically closer to (or further from) each other. This gives rise
to the idea of protected groups---for example, in guiding online search
completion or recommendation, female words might require protection in
that they should not be systematically closer to stereotypically female
job names, such as ``nurse'', ``librarian'', ``waitress'', and male
words require protection in that they should not be systematically
closer to toxic masculinity stereotypes, such as ``tough'', ``never
complaining'' or ``macho''.\footnote{However, for some research-related
  purposes, such as the study of stereotypes across history \citep{Garg2018years},
  embeddings which do not protect certain classes may also be useful.}

@@

One of the first measures of bias has been developed in \citep{Bolukbasi2016man}. The
general idea is that a certain topic is associated with a vector of real
numbers (the topic ``direction''), and the bias of a word is
investigated by considering the projection of its corresponding vector
on this direction. For instance, in \citep{Bolukbasi2016man}, the gender direction
\(\mathsf{gd}\) is obtained by taking the differences of the vectors
corresponding to ten different gendered pairs (such as
\(\overrightarrow{she} - \overrightarrow{he}\) or
\(\overrightarrow{girl} - \overrightarrow{boy}\)), and then identifying
their principal component.\footnote{Roughly, the principal component is
  the vector obtained by projecting the data points on their linear
  combination in a way that maximizes the variance of the projections.}
The gender bias of a word \(w\) is then understood as \(w\)'s projection
on the gender direction: \(\vec{w} \cdot gd\) (which, after normalizing
by dividing by \(\lVert w \rVert \,\lVert \mathsf{gd} \rVert\), is the
same as cosine similarity). Given a list \(N\) of supposedly gender
neutral words,\footnote{We follow the methodology used in the debate in
  assuming that there is a class of words identified as more or less
  neutral, such as \emph{ballpark, eat, walk, sleep, table}, whose
  average similarity to the gender direction (or other protected words)
  is around 0. See our list in Appendix \ref{app:custom} and a brief
  discussion in Subsection \ref{subsec:interpretability}.} and the
gender direction \(\mathsf{gd}\), the direct gender bias is defined as
the average cosine similarity of the words in \(N\) from \(\mathsf{gd}\)
(\(c\) is a parameter determining how strict we want to be):
\begin{align*}
\mathsf{directBias_c(N,gd)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},\mathsf{gd})\vert^c}{\vert N \vert }
\end{align*} \normalsize

The use of projections in bias estimation has been criticized for
instance in \citep{Gonen2019lipstick}, where it is pointed out that while a higher average
similarity to the gender direction might be an indicator of bias with
respect to a given class of words, it is only one possible manifestation
of it, and reducing the cosine similarity to such a projection may not
be sufficient to eliminate bias. For instance, ``math'' and ``delicate''
might be equally similar to a pair of opposed explicitly gendered words
(\emph{she}, \emph{he}), while being closer to quite different
stereotypical attribute words (such as \emph{scientific} or
\emph{caring}). Further, it is observed in \citep{Gonen2019lipstick} that most word pairs
retain similarity under debiasing meant to minimize projection-based
bias.\footnote{In \citep{Bolukbasi2016man} another method that involves analogies and
  their evaluations by human users on Mechanical Turk is also used. We
  do not discuss this method in this paper, see its criticism in
  \citep{Nissim2020fair}.}

A measure of bias in word embeddings that does not proceed by
identifying bias directions (such as a gender vector), the Word
Embedding Association Test (\textsf{WEAT}), has been proposed in
\citep{Caliskan2017semanticsBiases}. The idea here is that the bias between two sets of target
words, \(X\) and \(Y\) (we call them protected words), should be
quantified in terms of the cosine similarity between the protected words
and attribute words coming from two sets of stereotype attribute words,
\(A\) and \(B\) (we will call them attributes). For instance, \(X\)
might be a set of male names, \(Y\) a set of female names, \(A\) might
contain stereotypically male-related, and \(B\) stereotypically
female-related career words. The association difference for a particular
word \(t\) (belonging to either \(X\) or \(Y\)) is:

@@

\textsf{WEAT} is inspired by the Implicit Association Test (IAT)
\citep{Nosek2002harvesting} used in psychology, and in some applications it uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. In \citep{Caliskan2017semanticsBiases} the authors argue that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. In \citep{Lauscher2019multidimensional} the methodology is extended
to a multilingual and cross-lingual setting, arguing that using
Euclidean distance instead of cosine similarity does not make much
difference, while the bias effects vary greatly across embedding
models.\footnote{Interestingly, with social media-text trained
  embeddings being less biased than those based on Wikipedia.} A similar
methodology is employed in \citep{Garg2018years}. The authors employ word embeddings
trained on corpora from different decades to study the shifts in various
biases through the century.\footnote{Strictly speaking, these authors
  use Euclidean distances and their differences, but the way they take
  averages and averages thereof is analogous, and so what we will have
  to say about pre-averaging leading to false confidence applies to this
  methodology as well.}

\begin{figure}

\begin{center}\includegraphics[width=0.55\linewidth]{figures/WEATexample5} \end{center}
\caption{A simple example of a two-class set-up. Two groups of protected words, $X$ and $Y$ with two stereotypical attribute sets. An example of \textsf{WEAT} calculations follows.}
\label{eq:WEATexample}
\end{figure}

@@

\textsf{WEAT} has been developed to investigate biases corresponding to
a pair of supposedly opposing stereotypes, so the question arises as to
how to generalize the measure to contexts in which biases with respect
to more than two stereotypical groups are to be measured. Such a
generalization can be found in \citep{Manzini2019blackToCriminal}. The authors introduce Mean
Average Cosine distance (\textsf{MAC}) as a measure of bias. Let
\(T = \{t_1, \dots, t_k\}\) be a set of protected words, and let each
\(A_j\in \mathcal{A}\) be a set of attributes stereotypically associated
with a protected word where \(\mathcal{A}\). For instance, when biases
related to religion are to be investigated, they use a dataset of the
format illustrated in Table \ref{tab:religionOriginal}. The measure is
defined as follows:

@@

\begin{figure}

\begin{center}\includegraphics[width=0.5\linewidth]{figures/MACexample} \end{center}
\caption{A small subset of the religion dataset.  To each protected word in $T$  there corresponds one class of stereotypical attributes typically associated with it (and other classes of stereotypical attributes associated with different protected words).}
\label{fig:MACexample}
\end{figure}

@@

Notably, the intuitive distinction between different attribute sets
plays no real role in the \textsf{MAC} calculations. Equally well one
could calculate the mean distance of \emph{muslim} to all the
predicates, mean distance of \emph{christian} to all the predicates,
the mean distance of \emph{jew} to all the predicates, and then to take the
mean of these three means.

@@

\begin{figure}

\begin{center}\includegraphics[width=0.7\linewidth]{figures/figcosineDistributions8d} \end{center}

@@

\begin{figure}

\begin{center}\includegraphics[width=0.7\linewidth]{figures/figcosineDistributions8df} \end{center}

@@

\begin{figure}

\begin{center}\includegraphics[width=0.7\linewidth]{figures/figcosineDistributions8dg} \end{center}
